Leaf traits drive differences in above-ground biomass distribution among woody plant functional types

### 2015-11-15 manuscript.Rnw
ms and suppinfo including figures and captions

% 200 word max.
\begin{enumerate}
\item The partitioning of biomass into leaves and stems is an important component of  global vegetation models (GVMs).
\item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants. The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT. 
\item The ratio of leaf mass to woody above-ground mass (MF/AS) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that MF/MS, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. 
\item We partitioned the variance of MF/MS into three component variables (AF/AS, MF/AF and MS/AS), and each of these four variables into PFT, climate and plant height. Plant height explained most variation in MF/MS and MS/AS (but not H or MF/AF). PFT explained much variation in MF/MS and MF/AF (but not AF/AS and only for larger plants in MS/AS). 
\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area. We argue that climate effects will largely be indirectly through plant height.
\item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
\end{enumerate}

@@

\begin{itemize}
\item Dataset covers climate space pretty well. 
\item MF and MS are non-linearly related to plant height (Fig 2), as expected. Also note stem mass increases with height more quickly than leaf mass.
\item Leaf/stem biomass distribution was very different between PFTs (Fig 3a), and proportional to LMA when estimated at a common height, so that leaf area ratio did not differ between PFTs. Consistent with variance partitioning: MF/MS mostly explained by H and PFT, but LAR not much explained by PFT.
\item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate. Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently.
\item We show that AF was proportional to AS, apart from larger trees. Almost no PFT effects.
\item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape. Across all data, the relationship between MS and AS was not related to PFT, but nearly proportional to height (as expected). For larger trees, MS at a given AS*H was clearly lower for evergreen gymnosperms, reflecting differences in wood density. Again no climate effects.
\item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
\end{itemize}

@@

\begin{itemize}
\item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
\item We found weak and inconsistent effects of climate on all variables studied, with some exception for LMA. However, climate affects plant height, so we do expect global patterns in biomass distribution (Klein 2015, Simard 2011). These results differ from Reich2014 but we don't know why.
\item Not surprising that LMA was related to climate, but this did not drive differences in biomass distribution. 
\item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories. This is an active area of research, but is yet to be extended to patterns in biomass distribution.
\item Leaf area was just about proportional to AS, except for the largest trees. This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise). Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these. 
\item Stem mass index: Chave biomass estimation. We find support for this, with MS/AS nearly proportional to height. Conifers diverged, for the subset of larger trees only. Not sure why these differences don't show up for small plants but variance was very high!
\item Some paragraph on future work.
\item Consequences for GVMs.
\end{itemize}

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
    \fi
    \caption{Relationships between above-ground woody biomass ($M_S$), basal stem area ($A_S$) and plant height (H). (a) $M_S$ as a function of $A_S$ for the entire dataset. Lines are generalized additive model fits. (b) Much additional variation in M_S can be explained by H. Shown is the ratio $M_S/A_S$ as a function of H, with generalized additive model fits. (c) Stem mass as a function of $A_S*H$ for a subset of the data in panels a and b, where the product of H and $A_S$ was larger than 0.01. For this subset of larger trees, stem mass was nearly proportional to $A_S*H$ (SMA fits). (d) From the data in panel c, the average ratio $M_S/(A_S*H$) differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
    \label{fig:stemmassindex}
\end{figure}

>>>>

% 200 word max.
\begin{enumerate}
\item The partitioning of biomass into leaves and stems is an important component of global vegetation models (GVMs).
\item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants. The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT and plant height. 
\item The ratio of leaf mass to woody above-ground mass (MF/AS) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that MF/MS, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. 
\item We partitioned the variance of MF/MS into three component variables (AF/AS, MF/AF and MS/AS), and each of these four variables into PFT, climate and plant height. Plant height explained most variation in MF/MS and MS/AS (but not H or MF/AF). PFT explained much variation in MF/MS and MF/AF (but not AF/AS and only for larger plants in MS/AS). 
\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area, at a given plant size. We argue that climate effects will largely be indirectly through effects of plant height.
\item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
\end{enumerate}

@@

\begin{itemize}
\item Dataset covers climate space pretty well. 
\item MF and MS are non-linearly related to plant height (Fig 2), as expected. Also note stem mass increases with height more quickly than leaf mass.
\item Leaf/stem biomass distribution was very different between PFTs (Fig 3a), and proportional to LMA when estimated at a common height, so that leaf area ratio did not differ between PFTs. Consistent with variance partitioning: MF/MS mostly explained by H and PFT, but LAR not much explained by PFT.
\item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate. Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently (Fig S4).
\item We show that AF was proportional to AS (with the exception of very large trees). AF/AS was nearly unrelated to PFT or plant height.
\item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape. Across all data, the relationship between MS and AS was not related to PFT, but nearly proportional to height (as expected). For larger trees, MS at a given AS*H was clearly lower for evergreen gymnosperms, reflecting differences in wood density. Again no climate effects.
\item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
\end{itemize}

@@

\begin{itemize}
\item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
\item We found weak and inconsistent effects of climate on all variables studied, with some exception for LMA. However, climate affects plant height, so we do expect global patterns in biomass distribution (Klein 2015, Simard 2011). These results differ from Reich2014 but we don't know why. 
\item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories. This is an active area of research, but is yet to be extended to patterns in biomass distribution.
\item Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
\item Leaf area was just about proportional to AS, except for the largest trees. This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise). Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these. 
\item Stem mass index: Chave biomass estimation. We find support for this, with MS/AS nearly proportional to height. Conifers diverged, for the subset of larger trees only. Not sure why these differences don't show up for small plants but variance was very high!
\item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
\item Consequences for GVMs.
\end{itemize}

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
    \fi
    \caption{Relationships between above-ground woody biomass ($M_S$), basal stem area ($A_S$) and plant height (H). (a) $M_S$ as a function of $A_S$ for the entire dataset. Lines are generalized additive model fits. (b) Much additional variation in $M_S$ can be explained by H. Shown is the ratio $M_S/A_S$ as a function of H, with generalized additive model fits. (c) Stem mass as a function of $A_S*H$ for a subset of the data in panels a and b, where the product of H and $A_S$ was larger than 0.01. For this subset of larger trees, stem mass was nearly proportional to $A_S*H$ (SMA fits). (d) From the data in panel c, the average ratio $M_S/(A_S*H)$ differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
    \label{fig:stemmassindex}
\end{figure}

### 2015-11-15 manuscript_suppinfo.Rnw
ms and suppinfo including figures and captions

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Leaf area ratio ($A_F/M_T$) by plant height (H) and PFT. Each symbol is an individual plant. Lines are generalized additive model fits using species/dataset as a random effect. Shaded areas are approximate 95\% confidence intervals.}
    \label{fig:figureS3}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
    \caption{\emph{May not go in manuscript, just for notes at the moment}. Estimation of basal stem diameter from diameter at breast height for small plants. Because 36\% of the dataset is for plants with a height less than 1.3m, we used basal stem area in all analyses. For a subset of the data, dbh was measured but not d.ba, even though the plants were only marginally taller than breast height. In this figure, black points are the measured ratio of d.ba/d.bh (for plants where both were measured), red the predicted ratio from the fitted nonlinear model. Below 1.8m, the variance of the ratio d.ba/d.bh increases drastically, so we excluded those plants from the analysis (unless d.ba was actually measured).}
    \label{fig:figureS4}
\end{figure}

\clearpage
\subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.  However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \citep{dekauwe2014}. Note that we use 'partitioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods in main text). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.

\subsubsection*{Constant allocation fractions}

Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.

\subsubsection*{Allometric constraints}

In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep. For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).

\subsubsection*{Optimality}

Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}. One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.

\subsubsection*{Climate effects}

Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs. In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}. For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water. Likewise, Friedlingstein et al. \citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients. It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.

Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass. For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported. In this case, allocation proceeds to leaves only. In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.

\bibliography{references}

>>>>

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Leaf area ratio ($A_F/M_T$) by plant height (H) and PFT. Each symbol is an individual plant. Lines are generalized additive model fits using species/dataset as a random effect. Shaded areas are approximate 95\% confidence intervals.}
    \label{fig:figureS2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
    \caption{\emph{May not go in manuscript, just for notes at the moment}. Estimation of basal stem diameter from diameter at breast height for small plants. Because 36\% of the dataset is for plants with a height less than 1.3m, we used basal stem area in all analyses. For a subset of the data, dbh was measured but not d.ba, even though the plants were only marginally taller than breast height. In this figure, black points are the measured ratio of d.ba/d.bh (for plants where both were measured), red the predicted ratio from the fitted nonlinear model. Below 1.8m, the variance of the ratio d.ba/d.bh increases drastically, so we excluded those plants from the analysis (unless d.ba was actually measured).}
    \label{fig:figureS3}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{Partial regression plots for leaf area per stem basal area (AF/AS) and leaf mass per area (MF/AF). These two variables showed the strongest dependency on climate of all variables studied (see Table). Results are for a linear mixed-effects model (using species-dataset as a random effect), with plant height (and squared height), MAT, MAP (and squared terms), PFT, and interactions between MAP, MAT and PFT. Non-significant terms were removed (for full model, see SI text - \emph{to be added}). Partial regression plots were constructed with the visreg package in R. Each point is a model prediction when all other terms in the model are held at their average value. Transparent areas are approximate 95\% confidence intervals.}
    \label{fig:figureS4}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
    \caption{Comparison of three methods for variance partitioning of the five studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction). \textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in R2 when the independent variable (i.e. PFT, climate or height) was excluded from a model. Note that a GAM does not include interactions. \textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in R2 when the independent variable (and all its interactions) were removed from the model. R2 was calculated following Nakagawa and Schielzeth (2013) using the MuMIn R package (excluding random effects). Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables. Both methods identified AF/AS and MF/AF as the only variables with a practically significant dependence on climate (variance component > 0.1), see Fig S4. \textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by Chevan and Sutherland (1991) as implemented in the hier.part R package. Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
    \label{fig:figureS5}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
    \caption{\emph{Should come earlier}. A reviewer asked to see MF and AF as a function of total above-ground biomass. This also shows that MF at given MT diverges between PFTs (except for very small plants), but largely overlaps for AF/MT. Exception is for very (very) large trees, but we only have two datasets there.}
    \label{fig:figureS6}
\end{figure}

% 
% \clearpage
% \subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}
% 
% 
% Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.  However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \citep{dekauwe2014}. Note that we use 'partitioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods in main text). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
% 
% \subsubsection*{Constant allocation fractions}
% 
% Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
% 
% 
% \subsubsection*{Allometric constraints}
% 
% In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep. For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
% 
% \subsubsection*{Optimality}
% 
% Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}. One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
% 
% \subsubsection*{Climate effects}
% 
% Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs. In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}. For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water. Likewise, Friedlingstein et al. \citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients. It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
% 
% Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass. For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported. In this case, allocation proceeds to leaves only. In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
% 
% 
% \bibliography{references}
